---
title: "05. Survival Analysis and Time-to-Event Outcomes"
execute:
  warning: false
  message: false
---
  
## Motivation
  
Time-to-event outcomes introduce a new layer of complexity that is absent in standard classification and regression.

In survival analysis:
  
- outcomes are partially observed
- follow-up times differ across individuals
- censoring is informative for evaluation
- many common ML workflows fail silently

These properties make survival analysis especially vulnerable to **evaluation errors and leakage**.


## Why survival analysis is different
  
In earlier tutorials, each observation contributed a fully observed outcome.

In survival settings, each observation consists of:
  
- a time-to-event
- an event indicator (event vs censoring)

Ignoring this structure leads to invalid evaluation.

In particular, treating survival outcomes as binary labels discards timing information and introduces bias.


## Leakage risks unique to survival analysis
  
Survival workflows introduce additional leakage pathways:
  
- using full follow-up information during preprocessing
- computing time-dependent features globally
- evaluating survival models without respecting censoring
- mixing risk sets across resampling splits

These errors often go undetected and produce optimistic results.

The principles from **C1–C3** apply here with greater force.


## Data
  
We use a classical survival dataset with right censoring.

```{r}
library(fastml)
library(censored)
library(survival)
library(dplyr)

data(lung, package = "survival")

surv_data <- lung %>%
  filter(!is.na(time), !is.na(status)) %>%
  mutate(
    status = ifelse(status == 2, 1, 0),  # 1 = event, 0 = censored
    sex    = factor(sex, labels = c("male", "female"))
  )

head(surv_data)
```

This dataset includes:

- survival time (`time`),
- event indicator (`status`),
- baseline clinical covariates.


## Defining a survival task

In fastml, survival analysis is specified explicitly.

```{r}
fit <- fastml(
  data       = surv_data,
  label      = c("time", "status"),
  algorithms = c("cox_ph", "survreg"),
  impute_method = "remove"
)
```

This declaration ensures that:

- censoring is respected,
- models are trained under survival-specific assumptions,
- evaluation uses survival-appropriate metrics.

## Guarded Resampling for Survival Outcomes

Under guarded resampling:

- preprocessing is learned within each training split,
- risk sets are isolated by fold,
- evaluation uses only information available at assessment time.

No observation contributes future information to training.

This is essential for valid survival evaluation.

## Survival-Specific Metrics

Survival models cannot be evaluated using accuracy or ROC AUC.

Instead, `fastml` reports metrics such as:

- concordance index (C-index),
- time-dependent performance summaries,
- integrated Brier score (when applicable).

```{r}
fit$resampling_results$`cox_ph (survival)`$aggregated
```

These metrics account for censoring and timing.

## Variability across folds

As with earlier tutorials, fold-level variability matters.

```{r}
fit$resampling_results$`cox_ph (survival)`$folds
```

Survival metrics often exhibit higher variability due to:

- censoring patterns,
- limited numbers of events,
- heterogeneous follow-up times.

This variability is informative, not pathological.

## What fastml Does Not Allow Here

Consistent with [**C4. What fastml Deliberately Does Not Allow**](C4-what-fastml-does-not-allow.qmd), `fastml` prevents users from:

- converting survival outcomes to binary labels,
- evaluating survival models with classification metrics,
- preprocessing time-to-event data globally,
- detaching evaluation from resampling.

These restrictions are necessary for validity.

## Interpretation Cautions

Survival metrics describe ranking or calibration under censoring, not absolute risk.

They should not be interpreted as:

- probabilities of survival at specific times,
- causal effects,
- guarantees of clinical utility.

Guarded resampling ensures valid evaluation, not clinical validity.

## Summary

- Survival outcomes require task-specific evaluation.  
- Censoring introduces new leakage risks.  
- Guarded resampling is essential in survival analysis.  
- Variability is expected and informative.  

`fastml` is designed to encourage survival-appropriate workflows by constraining how outcomes, preprocessing, resampling, and evaluation are coordinated under supported execution paths.

## What comes next

**[06. Regression and Continuous Outcomes](06-regression-and-continuous-outcomes.qmd)**  
Regression and continuous outcomes under guarded resampling.

